Dotted Suffix Trees\\A Structure for Approximate Text Indexing

### 2005-08-11 paper/construction.tex
Small fixes. Longer, more detailed explanations.

Note that these conditions are very similiar to the conditions for the root and both can be implemented as shown in Figure~\ref{algo:copy-subtree}. This is an expression of Lemma~\ref{lemma:suffix-error}.

@@

To add dot links at every node, we need to add the dot link at the root and then for every node call the procedure addDotLink. This is shown in Figure~\ref{algo:construct}. The \textit{foreach} loop can be implemented using Depth First Search on the tree. In fact, the procedure addDotLink only needs to be called for certain nodes (those who do not posess an incoming suffix link). However, there is no readily available way to access only these nodes and the procedure above does little extra work.

>>>>

These conditions are an expression of Lemma~\ref{lemma:suffix-error}.
Note that these conditions are very similiar to the conditions for the root and both can be implemented as shown in Figure~\ref{algo:copy-subtree}.

@@

To add dot links at every node, we need to add the dot link at the root and then for every node call the procedure addDotLink. This is shown in Figure~\ref{algo:construct-1}. The \textit{foreach} loop at line~\ref{algoline:construct-1:foreach} can be implemented using Depth First Search on the tree. In fact, the procedure addDotLink only needs to be called for certain nodes (those who do not posess an incoming suffix link). However, there is no readily available way to access only these nodes and the procedure above does little extra work.

### 2005-08-11 paper/introduction.tex
Small fixes. Longer, more detailed explanations.

Since their introduction~\cite{weiner}, suffix trees have been one of the methods of choice for text indexing. They allow one to solve the following problem: given a text which has been preprocessed, and a pattern, efficiently find the places in the text where this pattern occurs.  However, for many real-life problems this statement is too restrictive. In these problems, one is interested in finding places in the text where an approximate form of the pattern occurs. Several algorithms have been proposed to solve this problem. \cite{navarro:guided} presents a survey of the problem and existing approaches to solving it at the time that work was written. Recently there have been some further work in the field, notably~\cite{maass:text} which also contains an updated overview of methods for solving the problem in addition to presenting one itself.

In this work we present an approach which solves this problem based on extending of suffix trees. The main advantage of this approach is that both the search and the index size are \emph{alphabet independent}, although the indexing time is not.

>>>>

Since their introduction~\cite{weiner}, suffix trees have been one of the methods of choice for text indexing. Suffix trees solve the following problem: given a text which has been preprocessed, and a pattern, efficiently find the places in the text where this pattern occurs.  However, for many real-life problems this is too restrictive. In this class of problems, one is interested in finding places in the text where an approximate form of the pattern occurs. Several algorithms have been proposed to solve this problem. \cite{navarro:guided} presents a survey of the problem and existing approaches to solving it at the time that work was written. Recently there have been some further work in the field, notably~\cite{maass:text} which also contains an updated overview of methods for solving the problem in addition to presenting one itself.

In this work we present an approach which solves this problem based on an extention of suffix trees. The main advantage of this approach is that both the search and the index size are \emph{alphabet independent}, although the indexing time is not.

### 2005-08-11 paper/search.tex
Small fixes. Longer, more detailed explanations.

\begin{figure}
\includegraphics{figures/mississipi-nodes}
\caption{Suffix Tree for the string \putstring{mississipi}}%
\label{fig:mississipi-nodes}
\end{figure}

@@

This algorithm runs in time $O(m^{k+1})$.

>>>>

\begin{figure}
\includegraphics[width=\textwidth]{figures/mississipi-nodes}
\caption{Suffix Tree for the string \putstring{mississipi}}%
\label{fig:mississipi-nodes}
\end{figure}

@@

This algorithm runs in time $O(m^{k+1})$:

The recursive call at line~\ref{algoline:search:followDot} means that a dot link was followed. There are at most $O(m^k)$ places where a dot link can be followed while descending a string~\footnote{For $k=1$ each point in the string. For $k>1$, each path for $k-1$, can now have at most $m$ extra branches.}. Following one path in the tree takes $O(m)$ time. Therefore, the total time is $O(m^{k+1})$.

### 2005-08-11 paper/space.tex
Small fixes. Longer, more detailed explanations.

How much space does a k-error dotted tree take? Suffix Trees have a number of nodes proportional to the size of the string (which we will refer to as $n$). One-error dotted trees have more nodes. The worst case presented in Figure~\ref{fig:aaaa} is a string of the form $a^n = \underbrace{aaa\cdots}_{n \mathit{times }a}$ which generates a one-dotted tree taking $O(n^2)$ nodes. For a larger number of errors $k$, it is easy to see that this leads to $O(n^{k+1})$ nodes.

@@

@@

Sofar, we have not really obtained much since, in the worst case, $h=O(n)$. However, the average expected case, is $h=O(\log n)$~\cite{devroye:note,szpankowski:unexpected}.

>>>>

How much space does a k-error dotted tree take? Suffix Trees have a number of nodes proportional to the size of the string (which we will refer to as $n$). One-error dotted trees have more nodes. The worst case presented in Figure~\ref{fig:aaaa} is a string of the form $a^n = aaa\cdots$ which generates a one-dotted tree taking $O(n^2)$ nodes. For a larger number of errors $k$, it is easy to see that this leads to $O(n^{k+1})$ nodes.

We can do better, however, by considering the expected case. First we need a pair of auxiliar definitions:

@@

Note that the weight of a tree \emph{is not the same} as the weight at the root.

@@

Sofar, we have not really obtained much since, in the worst case, $h=O(n)$. However, the average expected case, is $h=O(\log n)$~\cite{devroye:note,szpankowski:unexpected}. This allows us to conclude that, in the average case, we will have $O(n\log^k n)$ nodes in our tree.

### 2005-08-11 paper/structure.tex
Small fixes. Longer, more detailed explanations.

Being at a node in the suffix tree corresponds to being in several points in the string at the same time. We now add to each node an extra edge which corresponds to moving one character forward in all these points.

@@

In this paper, we will be interested in substitution of Hamming distance:

@@

In a suffix tree, it is possible to define a suffix link for each internal node. Both Ukkonen's~\cite{ukkonen} and McCreight's~\cite{mccreight} algorithms are linear time algorithms for constructing a suffix tree which includes suffix links.

@@

\begin{definition}[1-error dotted Tree]
We define a \emph{1-error dotted tree} as the tree which add to each node in a suffix tree, a new edge labeled by \putstring{.} which points to its error tree. The edge labeled \putstring{.} shall be called a \emph{dot link}.
\end{definition}

\begin{figure}
\includegraphics{figures/mississipi-1}
\caption{1-error dotted tree for \putstring{mississipi}}%
\label{fig:mississipi-1}
\end{figure}

>>>>

To extend the suffix tree, we consider that a node in the suffix tree corresponds to several points in the string at the same time (those points where the string on the path to the node occurs). We now add to each node an extra edge which corresponds to moving one character forward in all these points.

@@

In this paper, we will be interested in substitution distance (aka Hamming distance):

@@

In a suffix tree, it is possible to define a suffix link for each internal node. Both Ukkonen's~\cite{ukkonen} and McCreight's~\cite{mccreight} algorithms are linear time algorithms for constructing a suffix tree which includes suffix links. The suffix links are shown in Figure~\ref{fig:mississipi-0} as dotted links.

@@

\begin{definition}[1-error dotted Tree]
We define a \emph{1-error dotted tree} as the tree which is formed by adding to each node in a suffix tree, a new edge labeled by \putstring{.} which points to its error tree. The edge labeled \putstring{.} shall be called a \emph{dot link}.
\end{definition}

\begin{figure} 
\includegraphics[width=\textwidth]{figures/mississipi-1.eps}
% mississipi-1.eps: 300dpi, width=6.08cm, height=1.63cm, bb=0 0 718 193
\caption{1-error dotted tree for \putstring{mississipi}}%
\label{fig:mississipi-1}
\end{figure}
